---
title: "eco602_In- Class Likelihoods"
author: "Timothy Nsubuga"
date: "10/21/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---
Learning Objectives
Calculating likelihoods
Estimating distribution parameters using maximum likelihood

Data Files
You’ll need bird.sta.csv and hab.sta.csv in your data subfolder.


```{r}
require(here)
```


```{r}
read.csv(here("data", "bird.sta.csv"))
```
```{r}
read.csv(here("data", "hab.sta.csv"))
```

What is likelihood?
Likelihood is a measure of the relative probability of observing a set of one or more observations given a proposed model and parameter values.

We’ll be calculating likelihoods using probability distributions and their parameter values in this activity.
Likelihood of a single observation is proportional to the value of the probabiltiy density/mass function.

Likelihood of multiple independent observations is proportional the product of individual likelihoods.


What is maximum likelihood?
Maximum likelihood (ML) is a criterion for selecting a model.

The likelihood of a collection of observations is measured as the sum of log-likelihoods of individual observations in a dataset.

ML selection procedures select the model parameter values that maximize the sum of log-likelihoods.

We’ve already met another selection, least squares.



Recall that in contrast to ML, in which we maximize our criterion, we minimize our criterion when we use least squares.

Calculating Likelihood
Likelihood is calculated from the probabiltiy density/mass functions.

Remember the d___() functions in R?


Likelinood of Two Observations
Let’s practice calculating likelihoods:

How can we calculate the likelihood for 2 observations?
Recall the general procedure for calculating likelihoods:
Propose a model or distribution.
Propose a set of model or distribution parameter values.
Calculate the likelihood for each observation.
Multiply the individual likelihoods to get the joint likelihood.
In practice, you calculate the sum of the log-likelihoods.
Here’s an example in R:

Let’s say I’ve been conducting counts of Wilson’s Warblers at study sites in Oregon.
I’ve been to 2 sites where I counted 2, and 6 birds.

```{r}
x_observed = c(2, 6)
print(x_observed)
```

I think I can model the population of birds at the study sites with a Poisson distribution.

I know that the Poisson distribution has a single parameter: λ.
I also know that the mean and standard deviation of a Poisson distribution are equal to λ
Choosing a model:

I think the count of 2 is unusually low, so I decide to propose a Poisson distribution with λ=4.5 as a model of the counts of Wilson’s Warblers on my study sites.

I can use dpois() to calculate the probability mass for the two counts given a Poisson distribution with λ=4.5

```{r}
dpois(x = 2, lambda = 4.5)
```

```{r}
dpois(x = 6, lambda = 4.5)
```

I know the likelihood of observing those particular counts together is the product of the individual likelihoods:

```{r}
dpois(x = 2, lambda = 4.5) * dpois(x = 6, lambda = 4.5)
```

I can take advantage of vectorization in R by storing the counts in a single vector:

```{r}
wiwa_counts = c(2, 6)
dpois(x = wiwa_counts, lambda = 4.5)
```

Product of Likelihoods
I can more easily calculate the product now:

```{r}
prod(dpois(x = wiwa_counts, lambda = 4.5))
```

Sum of Log-Likelihoods
And the sum of log-likelihoods like:

```{r}
sum(log(dpois(x = wiwa_counts, lambda = 4.5)))
```

Likelihood of Many Observations
Now let’s say I want to find the value of that maximizes the likelihood for the counts of Wilson’s Warblers.

I first need to load the bird data:

```{r}
dat_bird = read.csv(here::here("data", "bird.sta.csv"))
dat_habitat = read.csv(here::here("data", "hab.sta.csv"))
dat_all = merge(dat_bird, dat_habitat)
```

Numerical Data Exploration
Next I need to do some numerical and graphical data exploration.

I’ll start with a 5-number summary, and then plot a histogram.

The summary:

```{r}
summary(dat_all$WIWA)
```

Graphical Exploration: Histograms
Default histogram
Next, we’ll plot a histogram of census counts.

Here’s the histogram that R produces with default settings:


```{r}
hist(dat_all$WIWA)
```


It’s not a pretty histogram.

Histogram with custom breaks
Let’s try setting the breaks argument to 7. This will suggest to R that it should create 7 bins (corresponding to observations between 0 and 7 wrens):

```{r}
hist(dat_all$WIWA, breaks = 7)
```

NOTE: if you look closely, this histogram includes both the observations of zero and one in the first bin. We requested 7 bins, but it only plotted 6!

When you use a single number for the breaks argument, R interprets it as a suggestion, but it’s binning algorithm may decide to use a different number!

Histogram with custom breaks attempt 2
One more try: we need to un-group the zero and one counts so that we see them as distinct bars in the histogram. We used a single number for the breaks argument to tell to try to automatically figure out how to divide the counts into 7 bins.

You can use a vector for the breaks argument. R attempts to estimate bin breakpoints when breaks is a single number, but it will honor your input if you provide a vector of breakpoints:


```{r}
hist(dat_all$WIWA, breaks = 0:7)

```

Still not exactly what we want. It turns out that for the first bin, R includes both endpoints in its count. In this case it’s counting all of the 0- and 1-observations together.

Histogram with custom breaks attempt 3: success!
We can trick R into only counting the lower endpoint if we cleverly manipulate the sequence that we give to bins.

```{r}
0:7 - 0.5
```

If we write code that subtracts a single value (a scalar in linear algebra lingo) from a vector, R will subtract the number from each element in the vector and return the output.

```{r}
hist(dat_all$WIWA, breaks = 0:7 - .5)
```

This works because the data were discrete counts. It also looks nicer because the bars are now centered over the census counts.

This trick doesn’t work as well for continuous data.
Histograms with (discrete) count data
If we wanted to use code like this with data for which we didn’t know the maximum value ahead of time we could write:


```{r}
par(mfrow = c(1, 2))
dat = dat_all$WIWA
hist(dat, breaks = 0:(max(dat) + 1) - 0.5, main = "Histogram of\nWilson's Warbler counts")

dat = dat_all$GRJA
hist(dat, breaks = 0:(max(dat) + 1) - 0.5, main = "Histogram of\nGray Jay counts")
```

```{r}
sum(log(dpois(x = dat_all$WIWA, lambda = 1.0)))
```













Questions 

Q1: As a group, experiment with different values of λ to find a value that maximizes the sum of log-likelihoods of the hypothetical observations.
Hint: you only need to consider 3 decimal digits!

```{r}
wiwa_counts = c(2, 6)
dpois(x = wiwa_counts, lambda = 4)
sum(log(dpois(x = wiwa_counts, lambda = 4)))
```

In the walkthrough, I tried using a Poisson distribution with λ=1.0 to model the set of census counts of Wilson’s Warbler.

You will create models for the census count of Winter Wrens.

Q2: Find the λ value of a Poisson model that makes all of the observed the Winter Wren’s census counts most likely.

Plot a histogram of Winter Wren counts (Check the metadata file to find the abbreviation for Winter Wrens). Make sure you choose an appropriate number of bins for the plot.

As a group, you’ll be searching for the value of λ that maximizes the sum of log-likelihood. In other words, you’ll find the the model parameter that makes the values of observed wren census counts most likely.

Hint: you only need to consider 2 decimal places in your value of λ.
Include your histogram into your report.
What value for λ did you select?
Show the R code you used to calculate the Poisson log-likelihood for the vector of Winter Wren census counts.

```{r}
 sum(log(dpois(x = dat_all$WIWR, lambda = 1.5))) 

```

```{r}
hist(dat_all$WIWR,  breaks = 0:7 - .5)
```

