---
title: "In-Class Confidence Intervals"
author: "Timothy Nsubuga"
date: "10/28/2021"
output: 
html_document:
  toc:  TRUE
  toc_float: TRUE 
  
---
# Critical Values {.tabset  .tabset - pills}

## Critical Values

Critical values are used for construction of parametric CIs.

For a two-sided 95% CI, they are the x-values that correspond to the locations on the x-axis, beyond which 5% of the total probability density lies.

For example, a 95% CI constructed from a Standard Normal distribution has critical Z values of ±1.96:

I can use the qnorm() function to find the 0.025% and 0.975% quantiles of the standard normal:

```{r}
qnorm(c(0.025, 0.975))
```

## Q1 90% Standard Normal Critical Values 

Q1 (1 pt.): Calculate the critical z-values for a 90% CI of the standard normal distribution. Show the R-code you used to perform the calculation.

```{r}
qnorm(c(0.05, 0.95))
```

## Q2-4 95% T-distribution critical values
Recall the t-distribution with it’s parameter: the degrees of freedom.

Q2 (1 pt.): Consult the help entry for qt() and calculate the critical values for df = 10. Show the R-code you used to perform the calculation.

```{r}
qt(c(0.025,.975),df = 10)
```


### Q3 (2 pts.): How many degrees of freedom are required for the 0.025% lower critical value of a t-distribution to match the 0.025% lower critical z-value (from the standard normal) to within one decimal place? Show the R-code you used to perform the calculation.

```{r}
qt(c(0.025,.975),df = 61)
```

## Q4 (1 pt.): How many degrees of freedom are required for the 0.025% lower critical value of a t-distribution to match the 0.025% lower critical z-value (from the standard normal) to within two decimal places? Show the R-code you used to perform the calculation.

```{r}
qnorm(.025)

qt(c(0.025),df = 475)
```

## Q5-6 95% CI using t-dist
Recall the general procedure for constructing a CI?

Suppose you know that the sample standard deviation for a group of 50 measurements is 3.14. The mean value is 10.0.

### Q5 (2 pts.): What are the critical t-values you would need to know to construct a 95% CI on the mean?

```{r}
#df = degress of freedom = n-1

z_crit95 = qt(c(0.025,.975),df = 49)
z_crit95
```

Q6 (3 pts.): Construct the interval. Show the R-code you used to perform the calculation.
CI = sample mean ± CI radius(sample standard error multiplied by critical z-value)
```{r}

  
  #Calculate the sd of mean of x]
 my_sd =  3.14
  
  #Get the size of x
  size = 50
  
  #perform SSE of mean calculation
  
 sse_mean =  my_sd/sqrt(size)

```

```{r}
my_mean = 10

```

```{r}
ci_radius = sse_mean * z_crit95
ci_radius
```

```{r}
#upper ci
ci_95up = my_mean + ci_radius
ci_95up
```
```{r}
#lower ci
ci_95low = my_mean - ci_radius
ci_95low
```

