---
title: "In-Class Models 1"
author: "Timothy Nsubuga"
date: "11/2/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Learning Objectives
Practice 1- and 2-sample t-tests
Data
We’ll use the penguins dataset for these exercises.

# Chinstrap Penguins: Sex
For the first few questions, we’ll use just the chinstrap penguin data.
Here’s some R-code to create a dataset containing only the Adelie penguins:

```{r get penguin data}
require(palmerpenguins)
```


```{r use just Adelie data}
dat_ade = droplevels(subset(penguins, species == "Adelie"))
```

Let’s look at a histogram:

```{r histogram of Adelie body mass }
hist(dat_ade$body_mass_g, main = "Adelie Penguins: Body Mass", xlab = "body mass (g)")
```

That seems like a weird distribution… Let’s see if we can figure out what’s going on.

Here’s my version of some boxplots, conditioned on sex:

# Questions {.tabset .tabset-pills}

## Q1 

### (2 pts.): Create separate boxplots of body mass for male and female Adelie penguins. Your boxplots do not have to be in the same panel like mine. Show the R-code you used to make the plots.

```{r boxplots onditioned on sex}
boxplot(
  formula = body_mass_g ~sex, 
        data = dat_ade)
```


## One-smaple t-test Q2-Q5 : For practice, let’s do a one-sample t-test. Recall that by default, when you provide the t.test() function with a group of measurements, it performs a one-sample t-test of the null hypothesis that the mean of the the group is equal to zero.

### Q2 (2 pts.): Perform a one-sample t-test of the alternative hypothesis that female Adelie penguins have a body mass different from zero grams. Note that this is a very silly alternative hypothesis. Is this a one- or two-tailed test? Show your R-code.

## You’ll need to do some subsetting.
## Check the help entry for t.test(). Specifically, check out the mu and alternative arguments.

```{r 1-sample t test for female ade}
pen_fem =  droplevels(subset(dat_ade, sex == "female"))

t.test(pen_fem$body_mass_g, y = NULL, 

       	alternative = "greater", 

       	mu = 0,   

conf.level = 0.95) 
```


## Q3 (1 pt.): Describe your conclusions based on the p-value of the t-test.

Since the p value is so small (< 2.2e-16), we can conclude that this number is significant, and female body mass is likely greater than 0. The p-value is very very small, so we could say that we can reject the null hypothesis that true mass of the Adelie female penguins is 0 g. Based on our confidence interval of 95% we find that if the true body mass was 0 grams it would be very unlikely that we would find the value we found from our sample of female Adelie penguins. 

## Q4 (2 pts.): Now, conduct a slightly less silly test: perform a one-sample t-test of the null hypothesis that male Adelie penguins have a mean body mass greater than 4000 grams. Is this a one- or two-tailed test?
```{r 1- sample t -test for male ade}
pen_men =  droplevels(subset(dat_ade, sex == "male"))

t.test(pen_men$body_mass_g, y = NULL, 

       alternative = "greater", 

  	mu = 4000,  

  	conf.level = 0.95) 
```


`**This a 1-tailed test as I am comparing a predictor to a value.**

## Q5 (1 pt.): Describe your conclusions based on the p-value of the t-test.






## Two-sample t- test Q6-10

