---
title: "In-Class Probability Calculations and Confidence Intervals"
author: "Timothy Nsubuga"
date: "10/14/2021"
output: html_document
---


Question 1: Binomial parameters
If I wanted to use a binomial distribution to model my six forest plots, what values should I use for the two parameters of a binomial distribution?

n = 6 and p = 2/3 

```{r}
dbinom(x = 4, size = 6, p = 2/3)
```
Question 2: Binomial probability 1
Use dbinom to calculate the probability of observing birds in exactly four of the six patches. Include your R-code in your answer.

```{r}
dbinom(x = 4, size = 6, p = 2/3)
```
Question 3: Binomial probability 2
Now, suppose I did a survey and observed no birds in my plots. Use dbinom to calculate the probability of observing no presences.

```{r}
dbinom(x = 0, size = 6, p = 2/3)
```

Question 4: Binomial Probabilities 3
Back to the binomial scenario (bird presence/absence in 6 forest plots).

Now use pbinom to calculate the probability of observing four or fewer presences in the 6 plots. Show your R code.

```{r}
pbinom( q = 4, size = 6, p = 2/3)
```

Question 5: Binomial Probabilities 4
Now use pbinom and the law of total probability to calculate the probability of observing four or more presences in the 6 plots. Show your R code

#pbinom(3,6,2/3) gives us the probability of 3 or fewer presences, 1 - this prob. gives us the probability of 4 or more presences 
```{r}
1 - pbinom(3,6,2/3)
```
#dfunctions are probability density at x
#pfunctions are cumulative probability to the left of x
