---
title: 'DataCamp Assignment: R Basics'
author: "Timothy Nsubuga"
date: "11/11/2021"
output: html_document
html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

# Questions {.tabset .tabset-pills}

## Section 1 Variables: Questions 1 - 6

Create:

A variable a that contains the text of your first name.

```{r var containing name}
a <- 'Tim'
```


A variable b1 that contains the number 45.6

```{r var containing number}
b1 <- 45.6
```

A variable b2 that contains the text “45.6”

```{r var containing word}
b2 <- "45.6"
```

A variable c1 that contains the sequence of integers from 0 to 3

```{r var containing sequence of numbers}
c1 <- 0:3
```

Hint: use the class() function to help with the questions below

#### Q1(1pt): What type of data is contained in the variable a?
```{r data type of a}
class(a)
```

**a contains data of type character.**

### Q2(1pt): What type of data is contained in the variable b1?

```{r data type of b1}
class(b1)
```

**b contains data of type numeric.**

### Q3(1pt): What type of data is contained in the variable b2?

```{r data type of b2}
class(b2)
```

**b2 contains data of type character.**

### Q4(2pt): What happens when you add b1 and b2 and why?

```{r sum of b1 and b2}
# b1 + b2

```

**Adding b1 and b2 results in an error because + is a binary operator that sums numeric data types and b2 is not a numeric data type.**

### Q5(1pt): Are the variables b1 and c1 of the same type?
```{r data type of c1 to compare to b1}
class(c1)
```

**No, c1 contains data of type integer and b1 contains data of type numeric.**

### Q6(3pt): Explain what happens when you add b1 and c1. Consider both the number of elements in each variable and the data types.

```{r sum of b1 and c1}
b1 + c1
```

**The result is a vector of 45.6 46.6 47.6 48.6. Element-wise summation is performed such that b1 is recycled 3 times and is summed with each element of c1.**

## Section 2 Vectors: Questions 7 - 9

Create a vector called v1 that contains a sequence of integers from -2 to 2.

When you print the contents of v1, it should look like this:

[1] -2 -1  0  1  2

```{r vector containing -2 to 2}
v1 <- -2:2
v1
```

Now, use v1 to create a new vector called v2 whose elements are the elements of v1 multiplied by 3. It should look like this:

[1] -6 -3  0  3  6

```{r v2 is product of V1 and 3}
v2 <- v1*3
v2
```

Finally, calculate the sum of all the elements in v2.
 
```{r sum of all elements in v2}
sum(v2)
```
 
### Q7(1pt): Show the R code you used to create v1.
 
 
```{r vector containing -2 to 2 again}
v1 <- -2:2
v1
```

 
### Q8(1pt): Show the R code you used to create v2.


```{r v2 is product of V1 and 3 again}
v2 <- v1*3
v2
```


### Q9(1pt): Show the R code you used to calculate the sum of elements in v2.

 
```{r sum of all elements in v2 again}
sum(v2)
```
 

## Section 3: Matrices Questions 10 - 11

Do you remember the byrow argument to the matrix() function?

Create a vector called vec_4 whose elements are the integers from 1 to 12.

```{r vector containing 1 to 12}
vec_4 <- 1:12
vec_4
```

Create a matrix mat_1 from vec_4 that has three rows and four columns. The values in mat_1 should be sequentially increasing by row.

For example, the first row of mat_1 should contain the values 1, 2, 3, 4.

```{r matrix of vec_4 by row }
mat_1 = matrix(data = vec_4, nrow = 3, ncol = 4, byrow = TRUE)
mat_1
```

Create a matrix mat_2 from vec_4 that has three rows and four columns. The values in mat_2 should be sequentially increasing by column.

For example, the first column of mat_1 should contain the values 1, 2, 3.

```{r matrix of vec_4 by col }
mat_2 = matrix(data = vec_4, nrow = 3, ncol = 4, byrow = FALSE)
mat_2
```

### Q10(1 pt) Show the code you used to create mat_1

```{r matrix of vec_4 by row again}
mat_1 = matrix(data = vec_4, nrow = 3, ncol = 4, byrow = TRUE)
mat_1
```

### Q11(1 pt) Show the code you used to create mat_2


```{r matrix of vec_4 by col again}
mat_2 = matrix(data = vec_4, nrow = 3, ncol = 4, byrow = FALSE)
mat_2
```

## Section 3: Lists Questions 12 - 14

Create a list, named my_list_1 with following three elements:

first element is numeric: 5.2
second element is a string "five point two"
third element is a vector of all integers from 0 to 5 [how do you do this?]
Name the elements in my_list_1:

“two”
“one”
“three”
Make sure the elements in your list are in the order specified.

Hint: remember the subsetting operators [[]] and $?

```{r list containing specified named elements}

my_list_1 <-list(5.2,"five point two", 0:5)
my_list_1

names(my_list_1) <- c('two','one','three' )
```

### Q12(2 pts): Show the R code you used to create my_list_1.


```{r list containing specified named elements again}

my_list_1 <-list(5.2,"five point two", 0:5)
my_list_1

names(my_list_1) <- c('two','one','three' )
```

### Q13(1pt): Show the R code that would select third element of the list.
```{r get third element of my_list_1}
my_list_1[[3]]
```

### Q14(1pt): Show the R code that selects the list element with the name “one”. Note: there are at least two ways to do this!

```{r select value of element named "one"}

my_list_1[[2]]

my_list_1$one
```

## Section 4: Logical Tests and Subsetting Questions 15 - 16

Run the following code to build a vector called my_vec:

```{r create my_vec}
my_vec = rep(1:3, 5)
my_vec
```

Use the logical equality test operator == to create a vector, my_bool_vec, of Boolean values from my_vec.

my_bool_vec should be the same length as my_vec.
my_bool_vec should have TRUE values in the positions where my_vec has values of 3.
You can run the following code to check that you have the correct values in my_bool_vec:

```{r create my_bool_vec}
my_bool_vec <- my_vec == 3
my_bool_vec
```

```{r check if my_bool_vec has teh correct values}
data.frame(my_vec, my_bool_vec)
```

Use my_bool_vec to retrieve all of the elements of my_vec that have a value of 3.

Hint: Use the square bracket subsetting operator: [].

Your result should look like this:

 [1] 3 3 3 3 3
 
```{r get all elements of my_vec that have a value of 3}
my_vec[my_bool_vec == TRUE]

```
 
### Q15(3 pts.) Show the R code that you used to create my_bool_vec.


```{r create my_bool_vec again}
my_bool_vec <- my_vec == 3
my_bool_vec
```

### Q16(2 pts.) Show the R code that you used to subset my_vec using my_boolean_vec.

```{r get all elements of my_vec that have a value of 3 again}
my_vec[my_bool_vec == TRUE]

```
 