---
title: "Probability and Frequentist Concepts"
author: "Timothy Nsubuga"
date: "10/22/2021"
output: 
html_document:
     toc: true
     toc_float: true
     toc_depth: 1
     
---

# Learning Objectives

 Use the distribution functions in R.
Sampling distribution insight
Seeing Theory
You’ll be using the Seeing Theory site to practice help with graphical intuition of Frequentist concepts.

 It’s a fantastic site, and I encourage you to check out the parts that we won’t cover in this assignment.

# Distribution functions for discrete distributions
 R implements probability, cumulative probability, and quantile functions for many parametric distributions.

 The naming convention is to use the name, or an abbreviation, of the distribution preceded by a prefix. For example with the binomial distribution the functions are:

 dbinom: the probability mass function

 pbinom: the cumulative mass function

 qbinom: the quantile function

Remember that discrete distributions use the term mass rather than density.
Remember that each distribution has different parameters, and you’ll have to look at the documentation on how to specify them with the # probability functions.
For a discrete parametric distribution, the distribution functions allow you to ask the following questions:

 What is the probability that I observe a value of exactly x?: The probability mass function, e.g. dbinom()
 What is the probability that I observe a value of x or less?: The cumulative mass function, e.g. pbinom()
What is the probability that I observe a value of x or more?: The cumulative mass function, e.g. pbinom()
 What is the median or 50th percentile?: The quantile function, e.g. qbinom()
 What is the 90th percentile?: The quantile function, e.g. qbinom()

# Distribution functions for continuous distributions#
 The naming convention is the same as for discrete distributions. For example with the normal distribution the functions are:

dnorm: the probability density function

pnorm: the cumulative density function

qnorm: the quantile function

Remember that continuous distributions use the term density rather than mass.
Remember that each distribution has different parameters, and you’ll have to look at the documentation on how to specify them with the probability functions.
# For a continuous parametric distribution, the distribution functions allow you to ask the following questions:

 Is a value of 1.2 or 2.4 more likely? - The probability density function, e.g. dnorm()
 What is the probability that I observe a value between 1.2 and 2.4?: The cumulative density function, e.g. pnorm()
 What is the probability that I observe a value of 1.3 or more?: The cumulative density function, e.g. pnorm()
 What is the probability that I observe a value of 2.4 or less?: The cumulative density function, e.g. pnorm()
 What is the 20th percentile of fish lengths?: The quantile function, e.g. qnorm()

# The law of total probability  
 You can use R’s distribution functions to answer many probability questions, but sometimes you have to be creative and use them in combinations.

 The sum of all events in the sample space is 1.0.

Sometimes it’s easier to calculate the probability of the complement of an event, rather than the event’s probability directly.

 Suppose you have an event E

 The complement of an event Ec is just the rest of the sample space not occupied by E.
 Pr(E)+Pr(Ec)=1.0

# Questions {.tabset .tabset-pills}

## Question 1
Q1 (2 pts.): What is the probability of observing a count of exactly 3 successes 
in a binomial distribution with parameters n = 4 and p = 0.75?

Include your answer and the R code you used to find it.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q1_dbinom()}
dbinom(x = 3, size = 4, p = 0.75)
```
## Question 2
Q2 (2 pts.): What is the probability of observing a count of 3 successes or fewer
in a binomial distribution with parameters n = 4 and p = 0.75?

Include your answer and the R code you used to find it.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q2_pbinom}
pbinom(q = 3, size = 4, p = 0.75)
```
## Question 3
Q3 (2 pts.): What is the probability of observing more than 3 successes 
in a binomial distribution with parameters n = 5 and p = 0.75?

Include your answer and the R code you used to find it.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q3_pbinom}
1 - pbinom(q = 3, size = 5, p = 0.75 )
```
## Question 4
Q4 (2 pts.): - What is the probability of observing a value of less than 
1.2 from a normally-distributed population with mean = 2 and standard deviation = 2? (check this again)

Include your answer and the R code you used to find it.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q4_pnorm()}
pnorm(q = 1.2, mean = 2, sd = 2)
```
## Question 5
What is the probability of observing a value of greater than 1.2 from a normally-distributed
population with mean = 2 and standard deviation = 2?

Include your answer and the R code you used to find it.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q5_pnorm()}
1 - pnorm(q = 1.2, mean = 2, sd = 2)

```

## Question 6
Q6 (4 pts.): - What is the probability of observing a value between 1.2 and 3.2 
from a normally-distributed population with mean = 2 and standard deviation = 2?

Include both your answer and the R code you used.
Note: To receive full credit, you cannot use lower.tail = FALSE in your code.

```{r q6_pnorm()}
pnorm( q = 3.2, mean = 2, sd = 2) - pnorm(q = 1.2, mean = 2, sd = 2)
```
## Question 7
Q7 (2 pts.): Describe how the shape of the histogram changes as you continue to press the sample button.

Follow the link to the Central Limit Theorem section.

Sampling Distribution
Central Limit Theorem
Navigate to the probability theory page on Seeing Theory.

Follow the link to the Central Limit Theorem section.

Sampling Distribution
Choose a set of α and β parameters that result in a skewed beta distribution.
Set the sample size to 1 and the draws to 50.
Hit the sample button several times and observe the evolution of the histogram.

For Questions 7 - 10 I uased alpha = 1.28 and Beta = 5

As I continue to press the sample button, the histogram shape increasingly matches my skewed beta distribution 

## Question 8
Q8 (2 pts.): Describe how the shape of the histogram changes as you continue to press the sample button.

Choose a set of α and β parameters that result in a skewed beta distribution.
Set the sample size to 2 and the draws to 50.
Hit the sample button several times and observe the evolution of the histogram

As I continue to press the sample button, the histogram's tail shortens and it becomes more symmetrical 

## Question 9
Q9 (2 pts.): Describe how the shape of the histogram changes as you continue to press the sample button.

Choose a set of α and β parameters that result in a skewed beta distribution.
Set the sample size to 15 and the draws to 50.
Hit the sample button several times and observe the evolution of the histogram.


As I continue to press the sample button, the histogram's tail shortens and it becomes almost perfect symmetrical with 
2 values having the highest frequency, the tails are even shorter

## Question 10
Q10 (2 pts.): Why is there such a drastic change in the shape of the sampling distribution 
when you change the sample size from 1 to 2?

Increasing the sample sample size decreases the sample standard error, a measure of how a sample mean compares to the population mean. Our approximation becomes closer as we increase the sample size from 1 to 2.

## Question 11
Q11 (2 pts.): What are the two main factors that determine the width of the sampling distribution of the mean?

The sample size and standard standard error determine the width of the sampling distribution of the mean because they are used to calculate confidence 
intervals which correlate with our uncertainly in using our sampling distribution's infer about the population.

## Question 12
There are 25 possible words consisting of a single character in the Library. There are 25×25 = 25^2 = 625 possible 2-character words.

Q12 (2 pts.): How many 3-character words are possible?

There are 25x25x25 =   15625 

## Question 13
Q13 (2 pts.): How many books would the Library contain if you added one additional position to the book size? Express your answer in terms of B.

Given the properties of the books in the Library:

410 pages
40 rows per page
80 positions per row
There are 410×40×80=1,312,000 positions for characters in each book.

Since there are 25 characters in the Library’s character set, there are a total of 25^1,312,000 possible books.

That’s a very large number. It’s about 2×10^1,834,097. Imagine a 2 followed by almost 2 million zeroes! It certainly wouldn’t fit on Earth.

That’s such a large number that it’s easier to just think of it in symbols. Let’s define the variable B as the number of books in the Library of Babel.

Q13 (2 pts.): How many books would the Library contain if you added one additional position to the book size? Express your answer in terms of B.

There would be B = 25^410x40x81 which is approximately infinite 


