---
title: 'Final Project: R Reference'
author: "Timothy Nsubuga"
date: "12/2/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages

### libary()

**The function library() loads an R package (code, documentation, data).**

Here is an example using the built in Palmer Penguins package.

```{r}
library(palmerpenguins)
```

### require()

**The require function also loads and R package (necessary when working in an R markdown in which a new R session is created every time the document is run).**

Here is an example using the Here package.

```{r}
require(here)
```

### here()

**The here function returns any subdirectory's file within an r project regardless of what the working directory is.**

### read.csv()

**The read.csv function converts a .csv file into tabulated data, converting it into a data frame.**

Here is an example using both here() and read.csv() functions on 2021 ginkgo data.


```{r}
ginkgo = read.csv(here("data","ginkgo_data_2021.csv"))
```

## Data Strucutures 

### c()

**The function c() combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).**

* All of the elements must be of the same type.
* I can’t combine character and numeric types in the same call to c()

Here’s two examples using numeric and character data types:

```{r}
## Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

## Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the print() function.

Typing the name of the vector into the console prints the contents

```{r}
## Typing the name of the vector into the console prints the contents
num_vec
```

```{r}
## The print() function accomplishes the same task:
print(char_vec)
```

### length()

**The function length() returns the number of elements contained in a vector.**

Here’s two examples using numeric and character data types:

```{r}
## Create a vector of numbers:
num_vec2  = c(5, 10, 36, 37, 45)

## Create a vector of characters:
char_vec2 = c("T", "ginkgo", "data is pretty cool")

## Get length of num_vec2 and char_vec2

length(num_vec2)

length(char_vec2)
```

### matrix()

**The function matrix() creates a 2-dimensional array(default: by column) of 1 data type from a given vector, designated number of rows, and columns.**

Here’s two examples using numeric and character data types:

 Create a vector of numbers:
```{r}
num_vec3  = c(3, 10, 32, 37, 45)
```


Create a vector of characters:

```{r}
char_vec3 = c("T", "salamander", "data is pretty cool", "R helped me like coding")
```


 Get make a vector using  num_vec3 
```{r}
matrix(num_vec3, 3,2)

```

 Note the error, the length of the vector must be a multiple of the number of rows.

try again

```{r}
num_vec4  = c(3, 10, 32, 37, 45,100)

mat_1 = matrix(num_vec4, 3,2)
 
mat_1
```

 make a matrix by row from char_vec3

```{r}
mat_2 = matrix(char_vec3, 2,2, byrow = TRUE)

mat_2
```

### data.frame()

**The function data.frame() creates a data frame which is a data structure that can contain several data types.**

Here is an example using the Gentoo Penguin.

Create vectors of Palmer Penguin facts

```{r}
# vector of characters 
names = c("Adelie", "Chinstrap", "Gentoo")

# vector of numerics 
avg_height_in = c(27.6,28,30)

# vector of logicals
flight = c(FALSE,FALSE,FALSE)
```

create data frame from these facts

```{r}
pen_facts = data.frame(names, avg_height_in, flight)
```

check the data type of pen_facts using the class() function

```{r}
class(pen_facts)
```

### nrow()

**The function nrow() returns the number of rows present in a vector, matrix or data frame.**

Here’s an example using the ginkgo data.frame(a data type that can contain several data types).


```{r}
nrow(ginkgo)
```

### ncol()

**The function ncol() returns the number of columns present in a vector, matrix or data frame.**

Here’s an example using the ginkgo data.frame.

```{r}
ncol(ginkgo)
```

### dim()

**The function dim() returns the number of rows and columns present in a vector, matrix or data frame.**

Here’s an example using the ginkgo data.frame.

```{r}
dim(ginkgo)
```

## Subsetting 

Subsetting extracts specified elements of vectors, matrices or elements

Use the ginkgo data for these examples:

### $ Subset a data frame by name: select one of the columns in the ginkgo data

```{r}
ginkgo$max_width
```

#### [] Use subset by position to:

select first row of the ginkgo data

```{r}
ginkgo[1,]
```


select the element in row 2, column 3

```{r}
ginkgo[2,3]
```

select the 3rd column of the ginkgo data

```{r}
ginkgo[,3]
```

## subset() 

Use this function to retrieve all the data for Adelie penguins (in the species column) from the peuguins dataset.

```{r}
subset(penguins$species, penguins == "Adelie")
```

##  Numerical Data Exploration 

### summary()

**The summary() function returns summary statistics for a given data frame.**

Here is an example using the Palmer Penguins data set.

```{r}
summary(penguins)
```

### mean()

**The mean() function calculates the arithmetic average of a given  numeric vector, matrix and data frame.**

Here is an example using the flipper flipper length of the Palmer Penguins(remove the NA values to preform the calculation).

```{r}
mean(penguins$flipper_length_mm, na.rm = TRUE)
```

### sd()

**The sd() function calculates the standard deviation of a given numeric vector, matrix and data frame.**

Here is an example using the flipper flipper length of the Palmer Penguins(remove the NA values to preform the calculation).

```{r}
sd(penguins$flipper_length_mm, na.rm = TRUE)
```

## Graphical Data Exploration

#### Scatterplot:

Using the ginkgo data, create a scatter plot of max leaf depth (x) and max leaf width (y).

plot() required arguments:
col =
pch =
cex =
main =
xlab =
ylab =
xlim =
ylim =

```{r}
plot(x = ginkgo$max_depth, ginkgo$max_width, xlim = c(0,50), ylim = c(0,50), main = "Ginkgo Width as Predicted by Depth < 50 mm", xlab = "Max Leaf Depth (mm)", ylab = "Max Leaf Width (mm)", col = "green", pch = 1, cex = 1)
```

#### hist()

Create a histogram of penguin flipper lengths. Required arguments:
breaks =

```{r}
hist(penguins$flipper_length_mm, breaks = 10, main = "Palmer Penguins Flipper Lengths", xlab = "Flipper Length (mm)", col = "blue")
```

### boxplot()

You must include two examples using the ginkgo data:

a simple boxplot of ginkgo petiole lengths

```{r}
boxplot(ginkgo$petiole_length, main = "Ginkgo Petiole Lengths", xlab = "Petiole length (mm)")
```

conditional boxplot of one of the continuous variables conditioned on the seeds_present column.

```{r}
boxplot(formula =  max_depth ~ seeds_present, main = "Ginkgo Max Depth as Predicted by Seed Presence", xlab = "Seeds Present ",
  data = ginkgo)
```

Create a 4-panel figure of histograms, arranged in a 2 by 2 grid. You may use any data you like, but each histogram must be different and have appropriate titles and axes. - par() required arguments: - mrow =

```{r}
par(mfrow = c(2,2))

hist( ginkgo$max_width, main = "Ginkgo Max Width", xlab = " Max Width (mm)")

hist( ginkgo$max_depth, main = "Ginkgo Max Depth", xlab = " Max Depth (mm)")

hist( ginkgo$notch_depth, main = "Ginkgo Notch Depth", xlab = " Notch Depth (mm)")

hist( ginkgo$petiole_length, main = "Ginkgo Petiole Length", xlab = " Petiole Length (mm)")
```

## Distribution Functions

### dnorm()

**dnorm() is the probability density function for the normal distribution.**

Here is an example that test where 1.2 or 2.4 is more likely:

```{r}
dnorm(1.2)

dnorm(2.4)
```

1.2 is more likely. 

### pnorm()

**pnorm() is the cumlative density function for the normal distribution.**

Here is an example to determine the probability of observing a value of less than 1.2 from a normally-distributed population with mean = 2 and standard deviation = 2


```{r }
pnorm(q = 1.2, mean = 2, sd = 2)
```

### qnorm() 

**qnorm() is the Quantile function for the normal distribution that returns the value associated wiith a specific probability (quantile).**

Here is an example to determine the median of the standard normal distribution

```{r}
qnorm(0.50)
```


### dbinom()

**dbinom() is the probability denisty function for the binomial distribution.**

Here is an example that determines the probability of observing a count of exactly 3 successes 

```{r }
dbinom(x = 3, size = 4, p = 0.75)
```

### pbinom()

**pbinom() is the cumulative mass function for the binomial distribution.**

Here is an example of observing the probability of observing a count of 3 successes or fewer with  4 trials and probability of success for each trial  = 0.75

```{r}
pbinom(q = 3, size = 4, p = 0.75)
```

### qbinom() 

**qbinorm() is the Quantile function for the binomial distribution that returns the value associated with a specific position within data (quantile).**


Here is an example to determine the median (50th quantile) of the binomial distribution  with 4 trials  and a probability of success for each trial equal to 0.75.

```{r}
qbinom(p = 0.50, size = 4, prob = 0.75)
```
