---
title: 'Part 2: Data Analysis'
author: "Timothy Nsubuga"
date: "12/13/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Directions and Questions {.tabset .tabset-pills}

You’ll perform a complete data analysis on the Delomys species data. You can do your work in a RMarkdown document, or an R script.

```{r}
require(here)
delomys = read.csv(here("data","delomys.csv"))
```

## Data Exploration

#### Numerical Exploration

Create a code chunk that includes the following:

Use summary() on the body mass and body length data columns in the Delomys data set to display summary statistics.

```{r summary() on the body mass and body length}
summary(delomys$body_mass)

summary(delomys$body_length)

```

Perform a test of normality on the body mass and length columns. You can use shapiro.test()

```{r test of normality on the body mass and length columns}
shapiro.test(delomys$body_mass)

shapiro.test(delomys$body_length)
```

### Graphical Exploration

You can adjust the size of the plots on your rendered document using the following code chunk arguments:

fig.height= fig.width=

You can adjust the aspect ratio using fig.aspect=

The code chunk format with an option looks like:

The code chunk format with an option looks like:

```{r fig.width=10}


require(palmerpenguins)


plot(bill_length_mm ~ body_mass_g, data = penguins)

```

Using code chunks, create the following plots, which you’ll use to answer the report questions:

A scatterplot of body mass and body length

```{r scatterplot of body mass and body length}
plot(delomys$body_mass, delomys$body_length, main = "Delomys Body Length as Predicted by Body Mass", xlab = "body mass", ylab = "body length")
```

A histogram of body mass

```{r}
hist(delomys$body_mass,  main = "Histogram of Delomys Body Mass", xlab = "body mass")
```

A histogram of body length

```{r histogram of body length}
hist(delomys$body_length, main = "Histogram of Delomys Body Length ", xlab = "body length")
```

A conditional boxplot of body mass, conditioned on species (column binomial)

```{r conditional boxplot of body mass conditioned on species}
boxplot(formula = body_mass ~ binomial, 
        main = " Delomys Body Length Conditioned on Species ", 
        data = delomys)
```


A conditional boxplot of body mass, conditioned on sex 
```{r conditional boxplot of body mass conditioned on sex }
boxplot(formula = body_mass ~ sex, 
        main = " Delomys Body Length Conditioned on Sex", 
        data = delomys)
```

A conditional boxplot of body mass, conditioned on both species and sex

```{r fig.width=10}
boxplot(formula = body_mass ~ binomial * sex, 
        main = " Delomys Body Length Conditioned on Species and Sex", 
        data = delomys,
       names = c("female\nDelomys\ndorsalis", "male\nDelomys\ndorsalis",
                 
                 "female\nDelomys\nsublineatus", "male\nDelomys\nsublineatus")
)
```

## Q1-4

Answer the following in your report:

Q1 (2 pts.): Qualitatively describe the relationship between body mass and length.

Does the relationship seem linear, curved, nonexistent?


**There seems to be a curved relationship between body mass and length.**

Q2 (2 pts.): Qualitatively describe the shapes of the histograms.
Do the data appear normally-distributed? Explain why or why not.

**Yes, the shape of the histogram of body mass is largely symetric about values between 40 - 50 and the histogram of body length is largely symmetric about values between 100 - 125.**

Q3 (3 pts.): Using both the histograms and normality tests, do you think the (unconditioned) body masses and body length are normally-distributed?
Contrast your visual assessment of normality to the results of the numerical normality tests.

**No, the (unconditioned) body masses and body length are not  normally-distributed because the Shapiro tests yield p -values of  4.33e-05 and 2.2e-16 respectively which means we can reject the null hypothesis that the body masses and body length  were sampled from a normally-distributed population.**

Q4 (2 pts.): Examine the conditional boxplots. Describe any graphical evidence you see for body mass differences based on species and/or sex.

**Species Delomys dorsalis have a larger body mass than delomys sublineatus.  Males have a larger body mass than females. Female Delomys dorsalis have larger body mass than male Delomys dorsalis and female Delomys sublineatus have larger body mass than male Delomys sublineatus.**

## Model Building

We know that the normality assumption applies to the residual values after we fit a model.

Using a code chunk, fit 5 models using lm():

Model 1: simple linear regression body_length ~ body_mass

```{r simple linear regression body_length ~ body_mass}
fit1 = 
  
  lm( 
    formula = body_length ~ body_mass,
    data = delomys)
```

Model 2: 1-way ANOVA body_mass ~ sex

```{r 1-way ANOVA body_mass ~ sex}
fit2 = 
  
lm( 
    formula = body_mass ~ sex,
    data = delomys)
```

Model 3: 1-way ANOVA body_mass ~ binomial

```{r 1-way ANOVA body_mass ~ binomial}
fit3 = 
  
lm( 
    formula = body_mass ~ binomial,
    data = delomys)
```

Model 4: 2-way additive ANOVA body_mass ~ sex + binomial


```{r 2-way additive ANOVA body_mass ~ sex + binomial}

fit4 = 
  
lm( 
    formula = body_mass ~ sex + binomial,
    data = delomys)
```


Model 5: 2-way factorial ANOVA body_mass ~ sex * binomial


```{r 2-way factorial ANOVA body_mass ~ sex * binomial}
fit5 = 
  
lm( 
    formula = body_mass ~ sex * binomial,
    data = delomys)
```


The first model predicts body length as a function of body mass The other models use the categorical variables binomial and sex to predict body mass.

Save your model objects to variables called fit1, fit2, fit3, fit4, fit5.

## Model Diagnostics

Let’s check whether our models fulfill the assumption of normality of the residuals.

First, use a graphical approach: plot histograms of the model residuals.

You can retrieve the model residuals using the residuals() function. For example, I could get the residuals from the first model using residuals(fit1).

```{r residuals from the first model}
residuals(fit1)
```

```{r residuals from the second model}
residuals(fit2)
```

```{r residuals from the third model}
residuals(fit3)
```

```{r residuals from the fourth model}
residuals(fit4)

```

```{r residuals from the fifth model}
residuals(fit5)
```

Use a code chunk to create histograms of the residuals of each of the 5 models.

```{r histograms of the residuals of each of the 5 models}
par(mfrow = c(3,2))

hist(residuals(fit1), main = "Histogram of Model 1 Residuals", xlab = "Residuals")

hist(residuals(fit2), main = "Histogram of Model 2 Residuals", xlab = "Residuals")

hist(residuals(fit3), main = "Histogram of Model 3 Residuals", xlab = "Residuals")

hist(residuals(fit4), main = "Histogram of Model 4 Residuals", xlab = "Residuals")

hist(residuals(fit4), main = "Histogram of Model 5 Residuals", xlab = "Residuals")
```

Next, use shapiro.test() on each model to test the null hypothesis that the residuals are drawn from a normally-distributed population.


```{r shapiro test on the first model}
shapiro.test(residuals(fit1))
```

```{r shapiro test on the second model}
shapiro.test(residuals(fit2))
```

```{r shapiro test on the third model}
shapiro.test(residuals(fit3))
```

```{r shapiro test on the fourth model}
shapiro.test(residuals(fit4))

```

```{r shapiro test on the fifth model}
shapiro.test(residuals(fit5))
```

## Q5-6

Answer the following in your report:

Q5 (2 pts.): What do you conclude about residual normality based on the numerical and graphical diagnostics?

**The normality assumption applies to the residual values after we fit a model. The shapiro test for models 1-5  yields p-values of 2.2e-16, 0.0001541, 8.103e-05, 6.805e-05 and 6.816e-05. This means we can reject the null hypothesis that the models' residuals are normally-distributed.**
, 
Q6 (1 pt.): Are violations of the normality assumption equally severe for all the models?

**No, violations of the normality assumption are most severe for model 1,   least severe for model 2 with models 3,4 and 5 within this range.**

## Model Interpretation

You can use the following code within a code chunk to print out a nicely formatted model coefficient table:

knitr::kable(coef(summary(my_model_fit)))

where my_model_fit is the name of your fitted model object.

You can use similar syntax to print a nicely formatted ANOVA table: knitr::kable(anova(my_model_fit))

Check out the digits argument to control how many decimal digits are printed.

## Q 7-9

### Body Length

Print the model coefficient table using summary() and answer the following:

```{r}
knitr::kable(coef(summary(fit1)))
```

Q7 (2 pts.): What is the magnitude of the mass/length relationship?

**The magnitude of the mass/length relationship is 0.880.**

Q8 (2 pts.): What is the expected body length of  an animal that weighs 100g?

**The expected body length is 76.124 + 0.880*100 = 164.124 INSERT UNITS.**

Q9 (2 pts.): What is the expected body length of an animal that weighs 0g?

**The expected body length is 76.124 + 0.880*0 = 76.124 INSERT UNITS.**

## Q 10-13

### Body Mass: Model Coefficients

Print the model coefficient tables for each of the body mass model fits.

```{r}
knitr::kable(coef(summary(fit2)))
```

```{r}
knitr::kable(coef(summary(fit3)))
```

```{r}
knitr::kable(coef(summary(fit4)))
```

```{r}
knitr::kable(coef(summary(fit5)))
```


Answer the following:

Q10 (1 pt.): What is the base level for sex?

**The base level for sex is female.**

Q11 (1 pt.): What is the base level for binomial?

**The base level for binomial is Delomys dorsalis.**

Q12 (1 pt.): Which sex is heavier?

**Males are heavier than females.**

Q13 (1 pt.): Which species is heavier?

**Delomys dorsalis are heavier than binomial Delomys sublineatus.**

## Q 14-16

### Body Mass: ANOVA

Print the ANOVA tables for each of the body mass models.

```{r}
knitr::kable(anova(fit2))
```

```{r}
knitr::kable(anova(fit3))
```


```{r}
knitr::kable(anova(fit4))
```


```{r}
knitr::kable(anova(fit5))
```

Answer the following in your report:

Q14 (1 pt.): Are sex and species significant predictors for body mass?

**The sex and species predictors have p-values of 0.0001951 and 0, respectively which are less than 0.005 and so yes, they are significant.**

Q15 (1 pt.): Is there a significant interaction?

**No, the p-value for the interaction is  0.950 which is greater than 0.005.**

Q16 (2 pts.): How dos the significance (as measured by p-value) of either of the main effects (sex and species) differ between the single-predictor models, the additive model, and the interactive model?

**Single-predictor models and the additive model both  have significance as opposed to the interactive model which does not.**

## Model Comparison: Body Mass

You built four different models of body mass. How do you choose the best one?

One option is to choose the model with the lowest AIC. You can calculate AIC using the appropriately named AIC() function.

Create a code chunk that calculates the AIC values for each of the body mass models.

```{r}
AIC(fit2)
```

```{r}
AIC(fit3)
```


```{r}
AIC(fit4)
```


```{r}
AIC(fit5)
```

## Q17-18

Q17 (1 pt.): Which two models have the lowest AIC?

**Model 4 has the lowest AIC of 12896.73 followed by Model 5 which has a value of 12898.72.**

Q18 (4 pts.): Which of the two models with lowest AIC scores would you select?

Explain your decision based on model fit and the complexity/understanding tradeoff.

**I would choose model 4 because the shapiro test for it yields a p-value of  6.805e-05 which is larger than the p-value for the shapiro test for model 5, 6.816e-05. This means that model 4 does not violate the normality assumption as much as model 5 and is therefore a better model fit. Model 4 also has a lower AIC which means it has a better balance between being a complex enough model that can be used to describe the data while being understandable, than model 5.**